
\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{ACT 2026}
\usepackage{breakurl}
\usepackage{underscore}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{tikz-cd}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,positioning}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  breaklines=true,
  frame=single,
  columns=flexible
}

\title{Functorial Composition over Task Decomposition: \\ A Categorical Kernel for AI-Human Computation}
\author{Anonymous Author
\institute{mo:os Project}
\email{author@example.com}
}
\def\titlerunning{Functorial Composition over Task Decomposition}
\def\authorrunning{A. Author}

\begin{document}
\maketitle

\begin{abstract}
Multi-agent AI frameworks decompose tasks top-down, producing orchestrations
that lack formal structural guarantees. We present \texttt{mo:os}, a runtime
kernel realized as a symmetric monoidal category where computational entities
are objects and state transitions are exactly four natural transformations
(\textsc{Add}, \textsc{Link}, \textsc{Mutate}, \textsc{Unlink}) over a typed
graph. Applying Lawvere's functorial semantics, \texttt{mo:os} separates
syntax (a JSON-schema-validated ontology of 21 object kinds and 16 morphism
types) from semantics (a deterministic Go execution engine implementing a
catamorphism $\mathit{state}(t) = \mathit{fold}(\mathit{log}[0..t])$). By
restricting LLMs to emit typed morphisms rather than orchestration plans,
complex behavior emerges from bottom-up operadic composition rather than
unstructured decomposition. We formalize the Container Category, prove closure
under parallel and sequential wiring, present the Recursive Semantic Bridge as
a structure-preserving functor, and report implementation metrics from a
working kernel: 3{,}729 lines of Go, 65 nodes, 80 wires, 14 HTTP endpoints,
and deterministic replay from an append-only morphism log.
\end{abstract}

%% ─────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

The rapid adoption of large language models (LLMs) as orchestrators of
multi-agent systems has exposed a fundamental architectural tension: the
frameworks that decompose tasks---LangChain, CrewAI, AutoGen, and their
successors---rely on the LLM's embedding space to simultaneously serve as
computation engine, state router, and correctness oracle. This conflation
produces what we term the \emph{semantic bottleneck}: as task complexity grows,
recombination of sub-results depends on ad-hoc aggregation that is brittle to
hallucination and context-window overflow.

We propose an inversion of control. Rather than asking the LLM to
\emph{orchestrate}, we restrict it to the role of a \emph{Fuzzy Processing
Unit} (FPU) that emits candidate morphisms over a strictly typed categorical
state machine. The deterministic kernel accepts or rejects these morphisms
against a semantic registry before committing them to an append-only log. This
separation---syntax from semantics, in the sense of Lawvere
\cite{lawvere1963}---yields structural guarantees that top-down decomposition
cannot provide.

\texttt{mo:os} is an open-source (MIT) runtime kernel, implemented in Go,
that realizes this architecture. Its categorical foundation draws on Fong and
Spivak's compositional framework \cite{fong2018}, Spivak's operadic wiring
diagrams \cite{spivak2013,spivak2013operad}, and Yau's finite presentation
results for wiring diagram operads \cite{yau2018}. Unlike Catlab (scientific
computing) or Statebox (blockchain consensus) \cite{statebox2019}, \texttt{mo:os}
targets the operational gap between LLM inference and industrial graph
state management.

\paragraph{Contributions.}
\begin{enumerate}
  \item We define the \emph{Container Category} $\mathcal{C}$ and prove it is
        closed under parallel and sequential wiring (Section~\ref{sec:container}).
  \item We formalize four invariant natural transformations and the operadic
        validation algebra that constrains them (Section~\ref{sec:bottleneck}).
  \item We construct the \emph{Recursive Semantic Bridge} as a
        structure-preserving functor $F\colon \mathbf{Syn} \to \mathbf{Sem}$
        and show that hydration stages correspond to strata of realization
        (Section~\ref{sec:bridge}).
  \item We present \emph{System~3} DAG reasoning as an extension of
        LogicGraph \cite{logicgraph2025} (Section~\ref{sec:system3}).
  \item We report metrics from a working implementation: 3{,}729 lines of Go,
        deterministic catamorphic replay, and a 21-kind ontology validated by
        an operad registry (Section~\ref{sec:implementation}).
\end{enumerate}

%% ─────────────────────────────────────────────────────────────────
\section{Background: Functorial Semantics and Operadic Composition}
\label{sec:background}

\subsection{Lawvere's Functorial Semantics}

Lawvere \cite{lawvere1963} established that an algebraic theory $\mathbf{T}$
can be separated from its models by defining models as product-preserving
functors $M\colon \mathbf{T} \to \mathbf{Set}$. This separation of syntax
(the theory) from semantics (the model) is the conceptual foundation of
\texttt{mo:os}: the JSON ontology is our syntax category, and the Go kernel is
a model that evaluates it.

\subsection{Operadic Wiring Diagrams}

Spivak \cite{spivak2013} formalized wiring diagrams as an operad
$\mathcal{W}$ where operations describe how boxes with typed ports are
interconnected. Rupel and Spivak \cite{spivak2013operad} extended this to
temporal wiring diagrams with directed, length-bearing wires suitable for
modeling dynamic information flow. Yau \cite{yau2018} proved that
$\mathcal{W}$ admits a finite presentation with 8 generators and 28 relations.

In \texttt{mo:os}, the 16 morphism types of our ontology correspond to
generators of a sub-operad of $\mathcal{W}$, and the semantic registry
enforces the operadic relations: each wire must connect admissible port types
with correct directionality.

\subsection{Dynamic Operads and Agent Composition}

Shapiro et al.\ \cite{shapiro2022} introduced \emph{dynamic operads} via the
monoidal double category $\mathbf{Org}$, showing how organized systems adapt
their operadic structure in response to internal and external pressures. This
framework supports the view of \texttt{mo:os} as a dynamic categorical system
where the graph topology evolves through morphism application while preserving
algebraic invariants.

Bonchi, Pavlovic, and Sobocinski \cite{bonchi2017functorial} extended
functorial semantics from algebraic to relational theories, providing tools for
reasoning about agent interactions as structure-preserving maps between
relational networks---a pattern we exploit in the \texttt{mo:os} wiring
algebra.

%% ─────────────────────────────────────────────────────────────────
\section{The Container Category}
\label{sec:container}

\begin{definition}[Container Category]
The \emph{Container Category} $\mathcal{C}$ has:
\begin{itemize}
  \item \textbf{Objects}: typed nodes $n = (\mathit{urn}, \tau, \sigma, \pi)$
        where $\mathit{urn}$ is a unique identifier, $\tau \in \mathcal{T}$ is
        a type drawn from the ontology ($|\mathcal{T}| = 21$), $\sigma \in
        \{S_0, \ldots, S_4\}$ is a stratum, and $\pi$ is a payload.
  \item \textbf{Morphisms}: directed wires $w\colon n_1 \to n_2$ specified by
        the 4-tuple $(\mathit{src}, p_s, \mathit{tgt}, p_t)$ where $p_s, p_t$
        are typed ports.
  \item \textbf{Identity}: for each node $n$, the identity wire
        $\mathrm{id}_n$ is the trivial self-loop (elided in practice).
  \item \textbf{Composition}: wires compose when the target port of one matches
        the source port of the next, subject to the operadic constraints of
        the semantic registry.
\end{itemize}
\end{definition}

The graph state at time $t$ is the carrier of the catamorphism:
\begin{equation}
\label{eq:cata}
\mathit{state}(t) = \mathrm{fold}\bigl(\mathit{log}[0..t],\; \varnothing,\;
\mathit{eval}\bigr)
\end{equation}
where $\varnothing$ is the empty graph (initial object) and $\mathit{eval}$
applies a single morphism envelope to produce a new state.

\begin{definition}[Stratum Lattice]
The five strata form a partial order:
\[
S_0\ (\text{Authored}) \leq S_1\ (\text{Validated}) \leq S_2\
(\text{Materialized}) \leq S_3\ (\text{Evaluated}) \leq S_4\ (\text{Projected})
\]
Higher strata may depend on lower strata; lower strata must not depend on
higher strata for meaning. $S_4$ projections are never ground truth.
\end{definition}

\begin{proposition}[Closure under Wiring]
\label{prop:closure}
Let $G_1, G_2$ be valid graph states in $\mathcal{C}$. The parallel
composition $G_1 \otimes G_2$ (disjoint union of nodes and wires) and the
sequential composition $G_1 \circ G_2$ (wire matching on shared ports) both
yield valid graph states in $\mathcal{C}$, provided all wires satisfy the
operadic constraints of the semantic registry.
\end{proposition}

\begin{proof}[Proof sketch]
Parallel composition preserves validity because disjoint union introduces no
new port conflicts. Sequential composition requires that matched ports have
compatible types and directions, which is enforced by the registry's
\texttt{ValidateLink} function checking source port direction (``out''), target
port direction (``in''), and target type membership in the source port's
admissible target set.
\end{proof}

\subsection{Ownership as Full Subcategory}

The OWNS morphism ($\mathit{MOR01}$) induces a forest structure on
$\mathcal{C}$. For any actor $a$, the \emph{scope} of $a$ is the full
subcategory $\mathcal{C}_a \hookrightarrow \mathcal{C}$ consisting of all
nodes transitively reachable from $a$ via OWNS wires. This is computed by
breadth-first search under a read lock, yielding all internal wires---those
whose source and target both lie in $\mathcal{C}_a$.

\input{figures/bottleneck}

%% ─────────────────────────────────────────────────────────────────
\section{The Semantic Bottleneck and Invariant Morphisms}
\label{sec:bottleneck}

\subsection{The Semantic Bottleneck}

The prevailing paradigm for multi-agent orchestration relies on the LLM's
embedding space as both computation engine and state router. We formalize this
structural failure as the \emph{semantic bottleneck}: a system architecture
where semantic inference and operational state management share a single,
untyped channel. As task complexity grows, this channel saturates, producing
hallucinated execution paths and brittle error recovery.

\subsection{Four Invariant Natural Transformations}

\texttt{mo:os} resolves the bottleneck by restricting all state mutations to
exactly four natural transformations:

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
NT & Signature & Semantic & Guard \\
\midrule
\textsc{Add}    & $\varnothing \to \mathrm{Node}$ & Create node &
  Duplicate URN $\Rightarrow$ \texttt{ErrNodeExists} \\
\textsc{Link}   & $N \times N \to \mathrm{Wire}$ & Create wire &
  Operadic port validation \\
\textsc{Mutate} & $N \to N$ & Update payload &
  Version CAS + $S_0$ block \\
\textsc{Unlink} & $\mathrm{Wire} \to \varnothing$ & Remove wire &
  Wire not found $\Rightarrow$ error \\
\bottomrule
\end{tabular}
\end{center}

These four operations are \emph{complete}: any graph state reachable from the
empty graph can be expressed as a finite composition of \textsc{Add},
\textsc{Link}, \textsc{Mutate}, and \textsc{Unlink} envelopes. The LLM (as
FPU) proposes candidate envelopes; the kernel validates and commits them.

\subsection{Operadic Validation}

The semantic registry $\mathcal{R}$ is an operad algebra over the port
structure of $\mathcal{C}$. For each type $\tau \in \mathcal{T}$, the registry
specifies:
\begin{itemize}
  \item \textbf{Allowed strata}: the set of valid strata for instances of $\tau$.
  \item \textbf{Port specifications}: for each named port, its direction
        (in/out) and admissible target types.
  \item \textbf{Mutability}: whether instances admit \textsc{Mutate}.
\end{itemize}

A \textsc{Link} envelope $w\colon n_1 \xrightarrow{(p_s, p_t)} n_2$ is valid
if and only if: (1) $p_s$ exists on $\tau(n_1)$ with direction ``out'', (2)
$p_t$ exists on $\tau(n_2)$ with direction ``in'', and (3) $(\tau(n_2), p_t)$
is in the admissible target set of $p_s$. This three-part check implements the
operadic composition constraints of $\mathcal{W}$.

%% ─────────────────────────────────────────────────────────────────
\section{The Recursive Semantic Bridge}
\label{sec:bridge}

We formalize the \texttt{mo:os} architecture as a realization of Lawvere's
functorial semantics. The database graph serves as the abstract syntax tree
(AST) of our language.

\begin{definition}[Syntax Category]
$\mathbf{Syn}$ has objects = Container types (the 21 kinds of the ontology)
and morphisms = wiring diagrams (the 16 morphism types).
\end{definition}

\begin{definition}[Semantics Category]
$\mathbf{Sem}$ has objects = Go runtime states and morphisms = deterministic
kernel evaluations (the fold step of Equation~\ref{eq:cata}).
\end{definition}

The \emph{Recursive Semantic Bridge} is the structure-preserving functor:
\begin{equation}
F\colon \mathbf{Syn} \to \mathbf{Sem}
\end{equation}

\begin{equation}
\begin{tikzcd}
\mathbf{Syn} \arrow[r, "F"] \arrow[d, "\alpha"'] & \mathbf{Sem} \arrow[d, "F(\alpha)"] \\
\mathbf{Syn}' \arrow[r, "F"'] & \mathbf{Sem}'
\end{tikzcd}
\end{equation}

where $\alpha$ represents a morphism (state transition) in the syntax, and
$F(\alpha)$ is the corresponding Go kernel evaluation. The functor preserves
composition: applying two morphisms sequentially in $\mathbf{Syn}$ yields the
same result as applying their $F$-images sequentially in $\mathbf{Sem}$.

\input{figures/hydration}

\subsection{Hydration as Stratum Promotion}

The hydration pipeline maps directly onto the stratum lattice:

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Stage & Stratum & Kernel Operation \\
\midrule
Authored   & $S_0$ & JSON instance files on disk \\
Validated  & $S_1$ & Schema check (JSON Schema draft-07) \\
Materialized & $S_2$ & \texttt{Materialize}: instance $\to$ Program \\
Evaluated  & $S_3$ & \texttt{ApplyProgram}: fold into graph state \\
Projected  & $S_4$ & \texttt{ScopedSubgraph}: BFS lens \\
\bottomrule
\end{tabular}
\end{center}

The key invariant is that $S_4$ projections are \emph{never} ground truth:
they are views derived from the evaluated state, and the graph can always be
reconstructed from the morphism log via replay (Equation~\ref{eq:cata}).

\subsection{Harness Agnosticism}

Because the functor $F$ maps syntax to semantics uniformly, any external
tool---an LLM, a database, a UI framework---is merely an object in
$\mathbf{Syn}$ whose interaction with the graph is governed by $F$. This
achieves \emph{harness agnosticism}: the UI surface is a direct, consistent
projection of the categorical state, requiring no intermediary state-management
synchronization. Nested sub-workspaces are submodules in the categorical
composition, inheriting structural integrity via algebraic closure.

%% ─────────────────────────────────────────────────────────────────
\section{System~3: Multi-Path DAG Reasoning}
\label{sec:system3}

We extend LogicGraph \cite{logicgraph2025} to define System~3 reasoning over
the morphism space:

\begin{itemize}
  \item \textbf{System~1}: Intuitive token prediction (standard LLM inference).
  \item \textbf{System~2}: Linear chain-of-thought (sequential reasoning).
  \item \textbf{System~3}: DAG exploration over typed morphism space.
\end{itemize}

In System~3, the deterministic kernel spawns candidate sub-graphs where each
edge is a structure-preserving morphism validated against the operad registry.
By applying tree search heuristics (MCTS or beam search) over the morphism
space, \texttt{mo:os} simulates branching future states. Because each branch
is restricted to valid categorical syntax, the search space is automatically
pruned of structurally inconsistent outcomes.

This enables high-stakes reasoning---infrastructure deployment, data
migration, multi-step planning---where the error surface of a single LLM
prompt would be prohibitively large. The typed morphism space provides a
natural pruning function: invalid compositions are rejected before evaluation,
and the catamorphic replay guarantees that any accepted path can be audited.

%% ─────────────────────────────────────────────────────────────────
\section{Implementation}
\label{sec:implementation}

\input{figures/architecture}

\texttt{mo:os} is implemented in Go (1.22+) as a single-binary kernel with
zero external dependencies beyond the standard library. The architecture
separates pure computation from effectful state management:

\begin{center}
\begin{tabular}{@{}llp{5.5cm}@{}}
\toprule
Layer & Package & Responsibility \\
\midrule
Types      & \texttt{cat}       & Core types: Node, Wire, GraphState, Envelope, Program. 4 morphism types, 5 strata. \\
Pure fold  & \texttt{fold}      & Catamorphism: \texttt{Evaluate}, \texttt{Replay}. Zero IO imports. \\
Operad     & \texttt{operad}    & Semantic registry: 21 TypeSpecs with port/direction/target validation. \\
Effect shell & \texttt{shell}   & \texttt{Runtime}: RWMutex-guarded state, \texttt{Apply}, \texttt{SeedIfAbsent}. \\
Transport  & \texttt{transport} & 14 HTTP routes on port 8000. \\
Hydration  & \texttt{hydration} & \texttt{Materialize}: instance JSON $\to$ Program. \texttt{HydrateAll}: batch boot. \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Catamorphic Replay}

The morphism log is an append-only JSONL file. State reconstruction is a left
fold:

\begin{lstlisting}[language=Go,caption={Pure replay (simplified)}]
func Replay(log []Envelope) GraphState {
    state := NewGraphState() // empty
    for _, env := range log {
        state = Evaluate(state, env)
    }
    return state
}
\end{lstlisting}

The \texttt{fold} package imports no IO packages (\texttt{os}, \texttt{net},
\texttt{sync}); it operates purely on values. The effect shell
(\texttt{shell.Runtime}) wraps the pure fold with \texttt{sync.RWMutex} for
concurrent read access and exclusive write access.

\subsection{Idempotent Seeding}

Boot uses \texttt{SeedIfAbsent}, which absorbs \texttt{ErrNodeExists} and
\texttt{ErrWireExists}. This makes repeated boots idempotent: the first boot
creates 13 seed morphisms; subsequent boots replay the log and produce zero
new entries.

\subsection{Metrics}

\begin{center}
\begin{tabular}{@{}lr@{}}
\toprule
Metric & Value \\
\midrule
Go source lines & 3{,}729 \\
Ontology object kinds & 21 \\
Ontology morphism types & 16 \\
Invariant natural transformations & 4 \\
Strata & 5 \\
HTTP routes & 14 \\
Nodes (post-hydration) & 65 \\
Wires (post-hydration) & 80 \\
Morphism log depth & 145 \\
Test packages (all green) & 6 \\
External dependencies & 0 \\
\bottomrule
\end{tabular}
\end{center}

%% ─────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\paragraph{Categorical frameworks.}
Catlab \cite{catlab2024} provides Julia libraries for computational category
theory; Statebox \cite{statebox2019} applies Petri nets to blockchain
consensus; Lynch and Breiner \cite{lynch2023nested} implement nested wiring
diagrams in Semagrams. None target the operational gap between LLM inference
and runtime graph state management.

\paragraph{Multi-agent orchestration.}
LangChain, CrewAI, AutoGen, and LangGraph rely on LLM-driven task
decomposition. These frameworks lack formal structural guarantees: correctness
depends on the LLM maintaining coherent state across unbounded context windows.

\paragraph{Operadic composition.}
Spivak's wiring diagram operads \cite{spivak2013,spivak2013operad} and Yau's
finite presentation results \cite{yau2018} provide the mathematical foundation
for our port-typed composition. Shapiro et al.'s dynamic operads
\cite{shapiro2022} formalize how operadic structure evolves---a pattern
\texttt{mo:os} realizes through its mutable graph topology.

\paragraph{Functorial semantics for computation.}
Bonchi et al.\ \cite{bonchi2017functorial} extend functorial semantics to
relational theories. Gu and Zanasi \cite{gu2021functorial} apply functorial
semantics as a unifying framework for logic programming. \texttt{mo:os}
applies these ideas to AI-human computation, where the ``programs'' are
morphism envelopes emitted by both human actors and LLM agents.

\paragraph{Category theory for AI.}
Pan \cite{pan2024token} proposes a categorical framework at the token level of
transformer architectures. Gavranovi{\'c} et al.\ \cite{gavranovic2024categorical}
develop an algebraic theory of deep learning architectures using 2-categories
of parametric maps. These works formalize AI \emph{internals} categorically;
\texttt{mo:os} instead formalizes the \emph{interface} between AI agents and
shared state.

\paragraph{Neuro-symbolic reasoning.}
LogicGraph \cite{logicgraph2025} benchmarks multi-path reasoning in
high-dimensional search spaces. \texttt{mo:os}'s System~3 extends this by
embedding DAG reasoning within a categorically typed morphism space, providing
structural pruning guarantees absent in untyped search.

%% ─────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We have demonstrated that the transition from top-down task decomposition to
bottom-up functorial composition provides a rigorous foundation for
AI-human computation. By anchoring LLMs as coprocessors within a typed
categorical kernel, \texttt{mo:os} resolves the semantic bottleneck while
preserving the expressiveness needed for complex multi-agent workflows. The
working implementation---3{,}729 lines of Go with zero dependencies, 21 object
kinds validated by an operad registry, and deterministic catamorphic
replay---shows that mathematical rigor is not an obstacle to practical systems
but rather their most robust enabling technology.

Future work includes: (1) an MCP bridge \cite{mcp2024} exposing the kernel as
a Model Context Protocol server with SSE transport; (2) a benchmark functor
implementing the classifying functor $\mathcal{T} \to \mathbf{Met}$ from
provider types to a metric space of benchmark scores; and (3) quantitative
comparison of morphism-based error recovery rates against traditional LLM
orchestration frameworks.

\bibliographystyle{eptcs}
\bibliography{references}

\end{document}
